Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-05-18 paper/paper.qmd
Write ML methods

TODO mikropml R package [@topcuoglu_mikropml_2021]

Balanced precision TODO

@@

::: {#fig-flowchart fig-cap="**CDI severity definitions.**
A) Decision flow chart to define CDI cases as severe according to the Infectious Diseases Society of America (IDSA) based on lab values,
the occurence of complications due to any cause (All-cause),
and the occurence of disease-related complications confirmed as attributable to CDI with chart review (Attrib).
B) The proportion of severe CDI cases labelled according to each definition.
An additional 'Pragmatic' severity definition uses the Attributable definition when possible,
and falls back to the All-cause definition when chart review is not available.
"}
![](figures/flowchart_sankey.png)
:::

::: {#fig-performance fig-cap="**Performance of ML models.**
Area Under the Receiver-Operator Characteristic Curve (AUROC) for the test sets
and cross-validation folds of the training sets,
and the Area Under the Balanced Precision-Recall Curve (AUPRC) for the test sets.
Left: models were trained on the full dataset, with different numbers of samples available for each severity definition.
Right: models were trained on the intersection of samples with all labels available for each definition.
"}
![](figures/ml-performance.png)
:::

>>>>

Random forest models were used to examine whether OTU data collected on the day
of diagnosis could classify CDI cases as severe according to four different
definitons of severity.
We used the mikropml R package v1.5.0 [@topcuoglu_mikropml_2021]
for all steps of the machine learning analysis.
We randomly split the data into an 80% training and 20% test set and repeated
this 100 times, followed by training models with 5-fold cross-validation.
Model performance was calculated on the test set using the area under the
receiver-operator characteristic curve (AUROC) and the Area under the balanced
precision-recall curve (AUBPRC).
Permutation feature importance was then performed to determine which OTUs
contributed most to model performance.
We reported OTUs with a significant permutation test in at least 80 of the 100
models.

Since the severity labels are imbalanced with different frequencies of severity
for each definition, we calculated balanced precision, the precision expected if
the labels were balanced.
The balanced precision and Area under the balanced precision-recall curve (AUBPRC)
were calculated with Equations 1 and 7 from @wu_improved_2021.

@@

::: {#fig-flowchart fig-cap="**CDI severity definitions.**
A) Decision flow chart to define CDI cases as severe according to the Infectious Diseases Society of America (IDSA) based on lab values,
the occurence of complications due to any cause (All-cause),
and the occurence of disease-related complications confirmed as attributable to CDI with chart review (Attrib).
B) The proportion of severe CDI cases labelled according to each definition.
An additional 'Pragmatic' severity definition uses the Attributable definition when possible,
and falls back to the All-cause definition when chart review is not available.
<!-- TODO table (supplementary?) showing counts & frequency of positives-->
"}
![](figures/flowchart_sankey.png)
:::

::: {#fig-performance fig-cap="**Performance of ML models.**
Area under the receiver-operator characteristic curve (AUROC) for the test sets
and cross-validation folds of the training sets,
and the Area under the balanced precision-recall curve (AUBPRC) for the test sets.
Left: models were trained on the full dataset, with different numbers of samples available for each severity definition.
Right: models were trained on the intersection of samples with all labels available for each definition.
<!-- TODO add plots of AUROC and AUBPRC curves -->
"}
![](figures/ml-performance.png)
:::
